Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-10 ind_files/04_unification_discussion.tex
Minor phrasing changes

where $w(x|\phi)$ and $\tau(x|\phi)$ are functionals that respectively describe the Malthusian fitness and per-capita turnover rate of type $x \in \mathcal{T}$ (now a continuous variable) in a population $\phi$. We then saw that this equation yields some well-known frameworks of quantitative genetics in the infinite population ($K \to \infty$) limit, thus illustrating consistency with known theories. If the `intuitive' version of It\^{o}'s formula for $\mathcal{M}_K(\mathcal{T})$ valued stochastic processes (obtained by `taking the limit' $m \to \infty$ in the It\^{o}'s formula for $m$-dimensional stochastic processes) holds, then the exact same steps carried out in Appendices \ref{App_density_to_freq} and \ref{App_stoch_var_eqns} will `go through' essentially unchanged for quantitative traits and yield the equations obtained by simply taking $m \to \infty$ in \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} (with sums replaced by integrals and derivatives replaced by functional derivatives as appropriate) as the `fundamental theorems' for the evolution of quantitative traits. Indeed,~\cite{week_white_2021} have recently proposed exactly the It\^o formula we would need via a heuristic It\^o multiplication table for $L^2(\mathbb{R}, m)$ valued processes with a.s. finite Hilbert-Schmidt norm and have shown that their heuristics are equivalent to the rigorous infinite-dimensional stochastic calculus proposed by~\cite{da_prato_stochastic_2014} for more general Hilbert space valued processes\footnote{If this sentence sounds like arcane gibberish to you, don't worry too much about the details - The essence is that~\cite{week_white_2021} have proven the `correct' formula we need holds whenever $b(x|\nu), d(x|\nu)$ and $\mathcal{M}_K(\mathcal{T})$ together fulfill certain technical requirements. My heuristics here are only a `first step' and have focused on accessibility over mathematical propriety, and thus make no attempt to verify whether and when these technical requirements are satisfied.}. Furthermore, using these `spacetime white noise heuristics' together with a particular functional form for (weak) solutions to certain SPDEs,~\cite{week_white_2021} have arrived at precisely the infinite-dimensional version of equations \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} obtained by `taking the limit' $m \to \infty$ in \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} from the `SPDE side'\footnote{If the well-known equivalence between SDEs and Fokker-Planck equations (see section \ref{sec_math_background}) carries through to infinite-dimensions, it becomes an equivalence between SPDEs and functional Fokker-Planck equations.} under certain additional assumptions required to ensure existence/uniqueness of solutions to the relevant SPDEs. Equation \eqref{disc_functional_field_eqns} can thus be argued to be the (functional) Fokker-Planck view of the stochastic processes for the evolutionary ecology of quantitative traits recently studied in~\cite{week_white_2021}. Like~\cite{week_white_2021}, I believe that rigorously establishing the precise criteria needed for general existence and uniqueness of solutions to equations of the form \eqref{disc_functional_field_eqns} (or the SPDE versions presented in~\cite{week_white_2021}, which resemble our equation \eqref{functional_langevin}) presents an important mathematical direction for future work that may yield new insights into both the biology of quantitative traits and the mathematics of measure-valued stochastic processes.

More generally, the intimate relation between It\^o SDEs and Fokker-Planck equations has been extensively exploited in the applied mathematics literature~\citep{van_kampen_stochastic_1981,oksendal_stochastic_1998, gardiner_stochastic_2009} since different problems are more appropriately attacked from one side than the other. In the infinite-dimensional case, the functional Fokker-Planck equation \eqref{disc_functional_field_eqns} (the `physics' side) allows us to attack questions of phenotypic clustering and evolutionary branching via powerful tools such as Fourier analysis (\cite{rogers_demographic_2012, rogers_modes_2015}; See also Appendix \ref{App_examples} for a general model-independent pedagogical derivation) that would have been significantly harder if one only used SPDEs. On the other hand, some questions are much more easily handled using tools from the `math' side of SPDEs such as infinitesimal generators and the martingale problem (see~\cite{champagnat_unifying_2006} or~\cite{week_white_2021}) rather than being forced to work with functional Fokker-Planck equations. Rigorously establishing a relation between functional Fokker-Planck equations of the form \eqref{disc_functional_field_eqns} and SPDEs of the form \eqref{functional_langevin} (or those studied in~\cite{week_white_2021}) for measure-valued branching processes could thus prove very fruitful for developing more integrative eco-evolutionary theory since it allows us to seamlessly transition between two alternative views of the same object. Importantly, the formalism I develop here in terms of functional Fokker-Planck equations likely does \emph{not} carry over to the study of higher dimensional quantitative traits (or populations which vary in two or more one-dimensional quantitative traits) because these processes are routinely badly behaved in higher dimensions: In particular, a (smooth) probability density function $P(\phi,t)$ frequently does not even exist in higher dimensions if one has any biologically non-trivial features such as interactions between types (\cite{fleming_measure-valued_1979, walsh_introduction_1986}; Also see for example~\cite{evans_measure-valued_1994}), making equation \eqref{disc_functional_field_eqns} entirely meaningless. My (admittedly limited) understanding is that such processes are also rather difficult to study in two or more dimensions from the SPDE side for similar reasons --- one is forced to work with distribution valued (rather than function valued) solutions and thus needs to use a considerable amount of advanced functional analysis to make progress~\citep{walsh_introduction_1986,carmona_stochastic_1999,balan_gentle_2018}.

To the best of my knowledge, a general formulation of stochastic field equations for the population dynamics of quantitative traits from the functional Fokker-Planck side in the manner I have carried out here has not been done before. Similar equations have been formulated for some specific stochastic models of quantitative trait evolution~\citep{rogers_demographic_2012,rogers_modes_2015} and population ecology of size-structured communities~\citep{odwyer_integrative_2009}. Stochastic field equations are also known in mathematical neurobiology~\citep{buice_field-theoretic_2007,bressloff_stochastic_2010,coombes_neural_2014}, and have recently been proposed in a model of collective motion~\citep{o_laighleis_minimal_2018}, suggesting that the study and formalization of such equations by pure mathematicians could have wide-spread applications. As is, such field equations are not on very solid mathematical footing, and are currently primarily used by physicists working in areas such as statistical field theory and are largely attacked using ingenious (and often heuristic) arguments and tools that are not necessarily mathematically well understood~\citep{carmona_stochastic_1999}. Equation \eqref{disc_functional_field_eqns} opens up the study of quantitative trait dynamics in finite fluctuating populations to analysis using various powerful tools of physics and stochastic processes such as duality~\citep{greenman_duality_2020}, the Fock space representation~\citep{dodd_many-body_2009, del_razo_probabilistic_2022}, and the path integral formalism~\citep{doi_second_1976, peliti_path_1985, dodd_many-body_2009, chow_path_2015, weber_master_2017}. Equation \eqref{disc_functional_field_eqns} also provides a more accessible alternative formulation that does not (explicitly) rely on tools from relatively `advanced' mathematical fields such as martingale theory and measure-valued processes that are typically used to study the evolution of quantitative traits in finite, fluctuating populations~\citep{dawson_stochastic_1975,fleming_measure-valued_1979,ethier_markov_1986,champagnat_unifying_2006,etheridge_mathematical_2011,week_white_2021}. This is intended to make the models and ideas more accessible to theorists without formal training in mathematical areas such as measure theory and functional analysis.

>>>>

where $w(x|\phi)$ and $\tau(x|\phi)$ are functionals that respectively describe the Malthusian fitness and per-capita turnover rate of type $x \in \mathcal{T}$ (now a continuous variable) in a population $\phi$. We then saw that this equation yields some well-known frameworks of quantitative genetics in the infinite population ($K \to \infty$) limit, thus illustrating consistency with known theories. If the `intuitive' version of It\^{o}'s formula for $\mathcal{M}_K(\mathcal{T})$ valued stochastic processes (obtained by `taking the limit' $m \to \infty$ in the It\^{o}'s formula for $m$-dimensional stochastic processes) holds, then the exact same steps carried out in Appendices \ref{App_density_to_freq} and \ref{App_stoch_var_eqns} will `go through' essentially unchanged for quantitative traits and yield the equations obtained by simply taking $m \to \infty$ in \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} (with sums replaced by integrals and derivatives replaced by functional derivatives as appropriate) as the `fundamental theorems' for the evolution of quantitative traits. Indeed,~\cite{week_white_2021} have recently proposed exactly the It\^o formula we would need via a heuristic It\^o multiplication table for $L^2(\mathbb{R}, m)$ valued processes with a.s. finite Hilbert-Schmidt norm and have shown that their heuristics are equivalent to the rigorous infinite-dimensional stochastic calculus proposed by~\cite{da_prato_stochastic_2014} for more general Hilbert space valued processes\footnote{If this sentence sounds like technical gibberish to you, don't worry too much about the details - The essence is that~\cite{week_white_2021} have proven the `correct' formula we need holds whenever $b(x|\nu), d(x|\nu)$ and $\mathcal{M}_K(\mathcal{T})$ together fulfill certain technical requirements. My heuristics here are only a `first step' and have focused on accessibility over mathematical propriety, and thus make no attempt to verify whether and when these technical requirements are satisfied.}. Furthermore, using these `spacetime white noise heuristics' together with a particular functional form for (weak) solutions to certain SPDEs,~\cite{week_white_2021} have arrived at precisely the infinite-dimensional version of equations \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} obtained by `taking the limit' $m \to \infty$ in \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} from the `SPDE side' under certain additional assumptions required to ensure existence/uniqueness of solutions to the relevant SPDEs. Equation \eqref{disc_functional_field_eqns} can thus be argued to be the (functional) Fokker-Planck view of the stochastic processes for the evolutionary ecology of quantitative traits recently studied in~\cite{week_white_2021}. Like~\cite{week_white_2021}, I believe that rigorously establishing the precise criteria needed for general existence and uniqueness of solutions to equations of the form \eqref{disc_functional_field_eqns} (or the SPDE versions presented in~\cite{week_white_2021}, which resemble our equation \eqref{functional_langevin}) presents an important mathematical direction for future work that may yield new insights into both the biology of quantitative traits and the mathematics of measure-valued stochastic processes.

More generally, the well-known intimate relation between It\^o SDEs and Fokker-Planck equations (see section \ref{sec_math_background}) has been extensively exploited in the applied mathematics literature~\citep{van_kampen_stochastic_1981,oksendal_stochastic_1998, gardiner_stochastic_2009} since many problems are much more easily attacked from one side than the other. In the infinite-dimensional case, working with the `physics' side through functional Fokker-Planck equation \eqref{disc_functional_field_eqns} allows us to use tools with an essentially `differential equation' flavor such as spectral methods (\cite{rogers_demographic_2012, rogers_modes_2015}; See also Appendix \ref{App_examples} for a general model-independent pedagogical example for studying phenotypic clustering and evolutionary branching), slow manifold (`adiabatic') approximations~\citep{parsons_dimension_2017},and WKB style expansions~\citep{rogers_demographic_2012}. On the other hand, many biologically interesting questions are much easier to approach from the `mathematics'\footnote{The `mathematics' vs `physics' distinction is just a caricature and is not intended to be read into or thought about too deeply :)} side by working with branching processes/SPDEs and using tools that have an essentially `probabilistic' flavor such as infinitesimal generators (semigroup theory), weak/mild solutions, and the martingale problem (See for example~\cite{etheridge_mathematical_2011} or~\cite{week_white_2021}). Rigorously establishing a relation between functional Fokker-Planck equations of the form \eqref{disc_functional_field_eqns} and SPDEs of the form \eqref{functional_langevin} (or those studied in~\cite{week_white_2021}) for measure-valued branching processes could thus prove very fruitful for developing more integrative eco-evolutionary theory since it allows us to seamlessly transition between two alternative views of the same object. Importantly, the formalism I develop here in terms of functional Fokker-Planck equations likely does \emph{not} carry over to the study of higher dimensional quantitative traits (or populations which vary in two or more one-dimensional quantitative traits) because these processes are routinely badly behaved in higher dimensions: In particular, a (smooth) probability density function $P(\phi,t)$ frequently does not even exist in higher dimensions if one has any biologically non-trivial features such as interactions between types (\cite{fleming_measure-valued_1979, walsh_introduction_1986}; Also see for example~\cite{evans_measure-valued_1994}), making equation \eqref{disc_functional_field_eqns} entirely meaningless. My (admittedly limited) understanding is that such processes are also rather difficult to study in two or more dimensions from the SPDE side for similar reasons --- one is forced to work with distribution valued (rather than function valued) solutions and thus needs to use a considerable amount of advanced functional analysis to make progress~\citep{walsh_introduction_1986,carmona_stochastic_1999,balan_gentle_2018}.

To the best of my knowledge, a general formulation of stochastic field equations for the population dynamics of quantitative traits from the functional Fokker-Planck side in the manner I have carried out here has not been done before. Similar equations have been formulated for some specific stochastic models of quantitative trait evolution~\citep{rogers_demographic_2012,rogers_modes_2015} and population ecology of size-structured communities~\citep{odwyer_integrative_2009}. Stochastic field equations are also known in mathematical neurobiology~\citep{buice_field-theoretic_2007,bressloff_stochastic_2010,coombes_neural_2014}, and have recently been proposed in a model of collective motion~\citep{o_laighleis_minimal_2018}, suggesting that the study and formalization of such equations by pure mathematicians could have wide-spread applications. As is, such field equations are not on very solid mathematical footing, and are currently primarily used by physicists working in areas such as statistical field theory and are largely attacked using ingenious (and often heuristic) arguments and tools that are not necessarily mathematically well understood~\citep{carmona_stochastic_1999}. Equation \eqref{disc_functional_field_eqns} opens up the study of quantitative trait dynamics in finite fluctuating populations to analysis using some of these tools --- For example, a careful reading of the literature provides many hints for how the major ideas behind the Fock space representation~\citep{del_razo_probabilistic_2022}, perturbative `loop diagrams/expansions'~\citep{dodd_many-body_2009}, and the path integral (really a `functional integral') formalism~\citep{doi_second_1976, peliti_path_1985, chow_path_2015, weber_master_2017} can all be co-opted to attack the sort of stochastic processes I formulate as a model for evolutionary dynamics via equation \eqref{disc_functional_field_eqns}. Equation \eqref{disc_functional_field_eqns} also provides a more accessible alternative formulation that does not (explicitly) rely on tools from relatively `advanced' mathematical fields such as martingale theory and measure-valued processes that are typically used to study the evolution of quantitative traits in finite, fluctuating populations~\citep{dawson_stochastic_1975,fleming_measure-valued_1979,ethier_markov_1986,champagnat_unifying_2006,etheridge_mathematical_2011,week_white_2021}. This is intended to make the models and ideas more accessible to theorists without formal training in mathematical areas such as measure theory and functional analysis.
